---
title: "Observational Studies"
subtitle: "Lecture 23"
date: "2026-08-04"
format:
  revealjs:
    output-file: 23-causal-inference-observational-slides.html
    include-in-header:
      - text: |
          <style>
          .reveal .title-slide h1 { font-size: 1.8em !important; }
          .reveal .title-slide .subtitle { font-size: 0.95em !important; }
          .reveal .title-slide .date { font-size: 0.75em !important; }
          .reveal .slides section.medium { font-size: 0.8em; }
          </style>
  html:
    output-file: 23-causal-inference-observational.html
auto-stretch: false
---

```{r}
#| label: load-packages
#| message: false
#| echo: false
library(tidyverse)
library(tidymodels)
library(knitr)
library(MatchIt)
```

# Recap: where we left off

## Last lecture: RCTs {.medium}

::: incremental
-   We showed that **randomization** is incredibly powerful: it makes $Z \perp X$, so any difference in $Y$ between groups must be due to treatment

-   This gave us the license to say **causes**: "receiving VWG OIT causes an increase in max tolerated wheat protein dose"

-   Today: what happens when we **can't** randomize?
:::

## When randomization isn't possible {.medium}

::: incremental
-   **Ethical**: you cannot randomly assign people to smoke, to not receive healthcare, or to experience childhood poverty

-   **Practical**: policy changes happen at scale. You can't "randomize" a country's minimum wage law

-   **Historical**: many causal questions are about events that already happened (retrospective)

-   Most real-world causal questions are answered with **observational data**.
:::

# Observational Studies

## What is an Observational Study? {.medium}

::: incremental
-   In an **observational study**, treatment is not assigned by the researcher, Subjects choose, or are assigned by nature or circumstance

- Surveys (including the census)

- Electronic Health Records

- Most sources of data can be thought of an observational study! Just have to define what the treatment is

- One key advantage over randomized experiments is the ease in collecting data! Usually has a much bigger sample size. 

## The issues

-  $Z$ is **not** independent of $X$: people who receive treatment differ systematically from those who don't

- Changes in $Y$ not directly attributable to changes in $Z$

[create small example table with like all of the x's matching the treatment Z, maybe like something about the new odyssey movie] Y is corisol levels, Z is whether watched movie, X is 

:::

## A concrete example {.medium} [CHANGE]

::: incremental
-   Students who seek tutoring tend to be those who are **struggling** --- lower prior GPA

-   If we compare exam scores, tutored students may still score lower on average, **even if tutoring genuinely helps**

-   The confounder (prior ability) affects both who gets tutoring ($Z$) and the outcome ($Y$)

-   Solution: we need a way to make treated and control groups **comparable** on $X$, without running an experiment
:::

## Key takeaway

::: incremental
-   In an observational study, a naive difference in group means is **not** a valid causal estimate

-   We need to adjust for confounders $X$ to recover the treatment effect

-   The rest of today is about two strategies for doing that
:::

# Matching

## The idea of matching {.medium}

recall covariate balance in a randomized study:

randomization balanced the covariates so that the only possible thing that changed was the treatment

::: incremental
-   For each **treated** unit, find one (or more) **control** units that look similar on $X$

-   Compare outcomes **within matched pairs** --- differences now reflect treatment, not confounding

-   We are trying to **mimic** a randomized experiment by constructing comparable groups from observational data

-   After matching: treated and control units have similar $X$ distributions, so $Z \perp X$ holds approximately
:::

## The key assumption {.medium}

::: incremental
-   Matching only works if we have measured **all** important confounders

-   Formally: **conditional ignorability**

    $$Y(0), Y(1) \perp Z \mid X$$

-   Given $X$, treatment assignment is "as good as random"

-   Also called **no unmeasured confounders** --- a strong, untestable assumption

-   This is the observational equivalent of randomization: we can't guarantee it, we have to argue for it
:::

# The LaLonde Study

## Context {.smaller}

LaLonde (1986) evaluated the **National Supported Work (NSW)** program: a 1970s job training intervention for disadvantaged workers.

::: incremental
-   **Cohort**: Disadvantaged workers (long-term unemployed, ex-offenders, welfare recipients)

-   **Design**: The original NSW had a randomized arm, but LaLonde compared its estimates against observational comparison groups drawn from national surveys --- showing how badly naive methods can fail

-   **Treatment** ($Z = 1$): Participation in the NSW job training program

-   **Control** ($Z = 0$): Non-participants drawn from observational comparison groups

-   **Outcome** ($Y$): Earnings in 1978 (`re78`)

-   **Covariates** ($X$): Age, education, race, marital status, prior earnings (`re74`, `re75`)
:::

## Data {.smaller}

```{r}
#| echo: true
data("lalonde", package = "MatchIt")
glimpse(lalonde)
```

-   `treat`: 1 = job training, 0 = no training
-   `re78`: earnings in 1978 (outcome $Y$)
-   `age`, `educ`, `re74`, `re75`, ...: pre-treatment covariates ($X$)

## Naive estimate

```{r}
#| echo: true
lalonde |>
  group_by(treat) |>
  summarise(mean_earnings = mean(re78))
```

::: incremental
-   The naive difference suggests training increases earnings

-   But are the two groups actually comparable before treatment?
:::

## Your turn {.smaller}

::: task
Look at the balance table below. Which variables show the largest differences between treated and control groups? What does this tell us about the naive estimate?

```{r}
#| echo: true
lalonde |>
  group_by(treat) |>
  summarise(
    mean_age  = mean(age),
    mean_educ = mean(educ),
    mean_re74 = mean(re74),
    mean_re75 = mean(re75)
  )
```
:::

## What the imbalance tells us {.medium}

::: incremental
-   Treated individuals have substantially **lower** prior earnings (`re74`, `re75`) than controls

-   The control group (from national surveys) is richer and more educated on average

-   A naive comparison overstates the earnings gap --- the treated group is catching up from a lower baseline, not just benefiting from training

-   We need to adjust for these differences before drawing causal conclusions
:::

## Nearest-neighbor matching {.smaller}

```{r}
#| echo: true
m_out <- matchit(
  treat ~ age + educ + re74 + re75,
  data   = lalonde,
  method = "nearest"
)
summary(m_out, un = FALSE)
```

## Estimate on matched data {.smaller}

```{r}
#| echo: true
m_data <- match.data(m_out)

linear_reg() |>
  fit(re78 ~ treat, data = m_data) |>
  tidy()
```

::: incremental
-   After matching, treated and control units are much more comparable on $X$

-   This estimate is our best causal guess **under the ignorability assumption**

-   Compare to the naive estimate --- confounding matters!
:::

# Propensity scores

## The curse of dimensionality {.medium}

::: incremental
-   Covariate matching works well when $X$ has only a few variables

-   With many covariates, it becomes nearly impossible to find good matches --- the covariate space is too sparse

-   Example: match exactly on age, education, prior earnings, race, marital status, region, ...

-   The probability of finding a near-identical control unit vanishes as the number of covariates grows

-   We need a smarter approach
:::

## The propensity score {.medium}

::: incremental
-   Instead of matching on the full $X$ vector, collapse it to a **single number**

-   The **propensity score**:

    $$e(X) = P(Z = 1 \mid X)$$

-   The probability of receiving treatment, given covariates

-   **Rosenbaum & Rubin (1983)**: if $Y(0), Y(1) \perp Z \mid X$, then also $Y(0), Y(1) \perp Z \mid e(X)$

-   Matching on this one number is sufficient!
:::

## Key takeaway

::: incremental
-   The propensity score **reduces a high-dimensional matching problem to a one-dimensional one**

-   We estimate $e(X)$ with logistic regression, then match on the predicted probability

-   Balance on $e(X)$ implies (approximate) balance on all of $X$
:::

## Estimating the propensity score {.medium}

We estimate $e(X)$ via **logistic regression** (which we already know!):

$$\log\frac{P(Z=1 \mid X)}{1 - P(Z=1 \mid X)} = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p$$

::: incremental
-   Fit logistic regression of $Z$ on $X$

-   Predicted probabilities $\hat{e}(X_i)$ are the estimated propensity scores

-   Match treated and control units with similar $\hat{e}(X)$
:::

## The workflow

1.  Fit a logistic regression model: $Z \sim X_1 + \cdots + X_p$
2.  Extract predicted probabilities as propensity scores
3.  Match treated to control units with similar scores
4.  Check **balance** on $X$ in the matched sample
5.  Estimate the treatment effect on the matched data

## Step 1 & 2: fit model, get scores {.smaller}

```{r}
#| echo: true
ps_model <- glm(
  treat ~ age + educ + re74 + re75,
  data   = lalonde,
  family = binomial
)

lalonde <- lalonde |>
  mutate(ps = predict(ps_model, type = "response"))

lalonde |>
  group_by(treat) |>
  summarise(mean_ps = mean(ps), sd_ps = sd(ps))
```

## Visualize the scores {.smaller}

```{r}
#| echo: true
ggplot(
  lalonde,
  aes(x = ps, fill = factor(treat, labels = c("Control", "Treated")))
) +
  geom_density(alpha = 0.5) +
  labs(
    x     = "Propensity score",
    y     = "Density",
    fill  = "Group",
    title = "Distribution of propensity scores by group"
  )
```

## Step 3 & 4: match and check balance {.smaller}

```{r}
#| echo: true
ps_match <- matchit(
  treat ~ age + educ + re74 + re75,
  data     = lalonde,
  method   = "nearest",
  distance = "glm"
)

summary(ps_match, un = FALSE)
```

## Step 5: estimate the treatment effect {.smaller}

```{r}
#| echo: true
ps_data <- match.data(ps_match)

linear_reg() |>
  fit(re78 ~ treat, data = ps_data) |>
  tidy()
```

::: incremental
-   After propensity score matching, we have a valid causal estimate (under ignorability)

-   This scales to many covariates, unlike direct covariate matching

-   Always check balance after matching --- if balance is poor, improve the propensity score model
:::

## Limitations of observational studies {.medium}

::: incremental
-   **No unmeasured confounders**: the key assumption is untestable --- if an important confounder is missing from $X$, our estimate is still biased

-   **Bad controls**: we need to choose $X$ carefully --- only include pre-treatment variables, never variables caused by $Z$

-   **Generalizability**: matching typically estimates the treatment effect on the **treated** (ATT), not the full population

-   **Model dependence**: the propensity score model matters --- wrong specification leads to poor balance and biased estimates

-   RCTs are still the gold standard; observational methods are our best option **when experiments are impossible**
:::

## Key takeaways {.medium}

::: incremental
-   Observational studies require careful adjustment for confounders $X$

-   **Covariate matching**: find control units similar on $X$ directly --- works well with few covariates

-   **Propensity score matching**: match on $P(Z = 1 \mid X)$ --- scales to high dimensions

-   Both require **no unmeasured confounders** --- an assumption that must be argued for, not tested from data

-   When you see an observational study making a causal claim, ask: *what confounders might they have missed?*
:::
