---
title: "Foundations of Causal Inference: Observational Studies"
subtitle: "Lecture 23"
date: "2026-08-04"
format: 
  revealjs:
    output-file: 23-causal-inference-observational-slides.html
  html: 
    output-file: 23-causal-inference-observational.html
auto-stretch: false
---

```{r}
#| label: load-packages
#| message: false
#| echo: false
library(tidyverse)
library(tidymodels)
library(knitr)
library(MatchIt)
```

# Revisiting confounding

## What changes without randomization?

::: incremental
-   In an **observational study**, the researcher does not assign treatment --- subjects choose (or are assigned by nature)

-   This means $Z$ is **not** independent of $X$: people who receive treatment differ systematically from those who do not

-   The backdoor path $X \to Z \to Y$ is open, and the naive difference in means **conflates** the treatment effect with confounding

-   $E[\bar{Y}_{\text{treated}} - \bar{Y}_{\text{control}}] \neq \tau$
:::

## The core problem

::: incremental
-   Students who seek tutoring tend to be those who are struggling --- lower prior GPA

-   If we compare exam scores, tutored students may *still* score lower even if tutoring helps

-   Conversely, employees who get promoted might be higher performers to begin with --- making the promotion look more effective than it is

-   We need a way to make treated and control groups **comparable** on $X$ without running an experiment
:::

# Matching on covariates

## The idea of matching

::: incremental
-   For each **treated** unit, find one (or more) **control** units that look similar on $X$

-   Compare outcomes within matched pairs --- differences now reflect treatment, not confounding

-   We are trying to **mimic** a randomized experiment by constructing comparable groups from observational data
:::

## Formally

::: incremental
-   The key assumption: **conditional ignorability**

$$Y(0), Y(1) \perp Z \mid X$$

-   Given covariates $X$, treatment assignment is "as good as random"

-   This is also called **no unmeasured confounders** --- a strong, untestable assumption

-   If it holds, matching on $X$ recovers the causal effect
:::

# An example with matching

## The data

We use the **LaLonde dataset**: a study of a 1970s job training program.
Participants were not randomly assigned.

```{r}
#| echo: true
data("lalonde", package = "MatchIt")
glimpse(lalonde)
```

-   `treat`: 1 = job training, 0 = no training
-   `re78`: earnings in 1978 (outcome $Y$)
-   `age`, `educ`, `re74`, `re75`, etc.: covariates $X$

## Naive estimate (no matching)

```{r}
#| echo: true
lalonde |>
  group_by(treat) |>
  summarise(mean_earnings = mean(re78))
```

::: incremental
-   The naive difference suggests training *increases* earnings

-   But treated and control groups may differ on pre-treatment characteristics --- we can't trust this yet
:::

## Check for imbalance

```{r}
#| echo: true
lalonde |>
  group_by(treat) |>
  summarise(mean_age  = mean(age),
            mean_educ = mean(educ),
            mean_re74 = mean(re74),
            mean_re75 = mean(re75))
```

::: incremental
-   Treated individuals have lower prior earnings (`re74`, `re75`) --- confounding is present

-   We need to adjust for this before drawing causal conclusions
:::

## Nearest-neighbor matching

```{r}
#| echo: true
m_out <- matchit(treat ~ age + educ + re74 + re75,
                 data   = lalonde,
                 method = "nearest")
summary(m_out, un = FALSE)
```

## Estimate on matched data

```{r}
#| echo: true
m_data <- match.data(m_out)

lm(re78 ~ treat, data = m_data, weights = weights) |> tidy()
```

::: incremental
-   After matching, the estimate is our best causal estimate under the ignorability assumption

-   Compare this to the naive estimate --- confounding matters!
:::

# The high-dimension problem

## Curse of dimensionality

::: incremental
-   Matching works well when $X$ has only a few variables

-   With many covariates, it becomes nearly impossible to find good matches --- the space is too sparse

-   Example: match exactly on age (35), education (12 years), prior earnings (\$8,000), race, marital status...

-   The probability of finding a near-identical control unit vanishes as the dimension of $X$ grows
:::

## We need a smarter approach

::: incremental
-   Instead of matching on the full $X$ vector, collapse it to a **single number**

-   The **propensity score**:

$$e(X) = P(Z = 1 \mid X)$$

-   The probability of receiving treatment, given covariates

-   **Rosenbaum & Rubin (1983)**: if $Y(0), Y(1) \perp Z \mid X$, then also $Y(0), Y(1) \perp Z \mid e(X)$

-   Matching on this one number is sufficient!
:::

# Propensity score matching

## Estimating the propensity score

We estimate $e(X)$ via **logistic regression**:

$$\log\frac{P(Z=1 \mid X)}{1 - P(Z=1 \mid X)} = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p$$

::: incremental
-   Fit logistic regression of $Z$ on $X$

-   Predicted probabilities $\hat{e}(X_i)$ are the estimated propensity scores

-   Match treated and control units with similar $\hat{e}(X)$ values
:::

## The workflow

1.  Fit a logistic regression model: $Z \sim X_1 + \cdots + X_p$
2.  Extract predicted probabilities as propensity scores
3.  Match treated to control units with similar scores
4.  Check **balance** on $X$ in the matched sample
5.  Estimate the treatment effect on the matched data

# Propensity score matching example

## Step 1 & 2: fit model, get scores

```{r}
#| echo: true
ps_model <- glm(treat ~ age + educ + re74 + re75,
                data   = lalonde,
                family = binomial)

lalonde <- lalonde |>
  mutate(ps = predict(ps_model, type = "response"))

lalonde |>
  group_by(treat) |>
  summarise(mean_ps = mean(ps), sd_ps = sd(ps))
```

## Visualize the scores

```{r}
#| echo: false
ggplot(lalonde, aes(x = ps, fill = factor(treat))) +
  geom_density(alpha = 0.5) +
  labs(x = "Propensity Score",
       y = "Density",
       fill = "Treatment",
       title = "Distribution of Propensity Scores by Group") +
  scale_fill_manual(values = c("steelblue", "tomato"),
                    labels = c("Control", "Treated"))
```

## Step 3 & 4: match and check balance

```{r}
#| echo: true
ps_match <- matchit(treat ~ age + educ + re74 + re75,
                    data     = lalonde,
                    method   = "nearest",
                    distance = "glm")

summary(ps_match, un = FALSE)
```

## Step 5: estimate the effect

```{r}
#| echo: true
ps_data <- match.data(ps_match)

lm(re78 ~ treat, data = ps_data, weights = weights) |> tidy()
```

::: incremental
-   After propensity score matching, we have a valid causal estimate (under ignorability)

-   The propensity score approach scales to many covariates, unlike direct matching

-   Always check balance after matching --- if it's poor, consider a different model for the propensity score
:::

## Key takeaways

::: incremental
-   Observational studies require careful adjustment for confounders $X$

-   Matching reconstructs comparability between treated and control groups

-   **Direct matching** works with few covariates; **propensity score matching** scales to high dimensions

-   All methods require **no unmeasured confounders** --- this assumption cannot be tested from data alone

-   Sensitivity analyses can assess how robust conclusions are to potential unmeasured confounding
:::
