---
title: "AE 17: Diving back into Paris 2024"
format:
  html:
    self-contained: true
---

In this application exercise, we will use bootstrapping to construct confidence intervals.

## Packages

We will use **tidyverse** and **tidymodels** for data exploration and modeling and the **knitr** package for formatting tables. The dataset was scraped from Wikipedia and the Olympics website.

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
library(tidymodels)
library(openintro)
library(knitr)
```

## Data

The data represent the scores of every dive during the final rounds of the 2024 Paris Olympic games diving events. Each diver performs six dives. Suppose we are interested in how well we can predict a diver's last round dive score from their running total of five dives (i.e. is there a "streak" in diving?)

```{r}
#| label: glimpse-data
diving_raw <- read_csv("data/diving.csv")
diving <- diving_raw |>
  mutate(dive_number = str_c("Dive_", dive_number)) |>
  select(athlete, country, event, dive_number, points) |>
  pivot_wider(
    names_from = dive_number,
    values_from = points
  ) |>
  mutate(running_total = Dive_1 + Dive_2 + Dive_3 + Dive_4 + Dive_5)

glimpse(diving)
```

## Model

```{r}
#| label: fit-model
#| message: false
#| warning: false
diving_fit <- linear_reg() |>
  fit(Dive_6 ~ running_total, data = diving)

tidy(diving_fit) |>
  kable(digits = 2)
```

## Bootstrap confidence interval

### 1. Calculate the observed fit (slope)

```{r}
#| label: observed-fit
# add code here
```

### 2. Take *n* bootstrap samples and fit models to each one.

Fill in the code, then set `eval: true`.

```{r}
#| label: bootstrap
#| eval: false
n = 100
set.seed(24601)

boot_fits <- ______ |>
  specify(______) |>
  generate(reps = ____, type = "bootstrap") |>
  fit()

boot_fits
```

-   Why do we set a seed before taking the bootstrap samples?

-   Make a histogram of the bootstrap samples to visualize the bootstrap distribution.

```{r}
#| label: boot-hist
# add code here
```

### 3. Compute the 95% confidence interval as the middle 95% of the bootstrap distribution

Fill in the code, then set `eval: true`.

```{r}
#| label: calc-ci
#| eval: false
get_confidence_interval(
  boot_fits,
  point_estimate = _____,
  level = ____,
  type = "percentile"
)
```

## Changing confidence level

### Modify the code from Step 3 to create a 90% confidence interval.

```{r}
#| label: 90-ci
# add code here
```

### Modify the code from Step 3 to create a 99% confidence interval.

```{r}
#| label: 99-ci
# add code here
```

-   Which confidence level produces the most accurate confidence interval (90%, 95%, 99%)? Explain.

-   Which confidence level produces the most precise confidence interval (90%, 95%, 99%)? Explain.

-   If we want to be very certain that we capture the population parameter, should we use a wider or a narrower interval? What drawbacks are associated with using a wider interval?
