---
title: "AE 20: Wheat Oral Immunotherapy"
format:
  html:
    self-contained: true
---

Nowak-Węgrzyn et al. (2019) conducted a multicenter, randomized, double-blind, placebo-controlled clinical trial to evaluate whether vital wheat gluten (VWG) oral immunotherapy (OIT) is effective at increasing wheat tolerance in children and young adults with wheat allergy.
In today's AE, you will analyze a dataset simulated to represent this study using the tools from lecture.

(<https://pubmed.ncbi.nlm.nih.gov/30389226/>)

# Packages

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
library(tidymodels)
```

# Data

```{r}
#| message: false
set.seed(42)
n <- 46
wheat_trial <- tibble(
  id                 = 1:n,
  age                = round(runif(n, 4, 22), 1),
  baseline_wheat_ige = round(rexp(n, rate = 1/30), 1),
  vwg_oit            = c(rep(1, 23), rep(0, 23))
) |>
  mutate(
    max_dose_mg = pmax(0, round(
      150 + 950 * vwg_oit - 3 * baseline_wheat_ige + 12 * age + rnorm(n, sd = 90)
    ))
  )

glimpse(wheat_trial)
```

Variables:

| Variable             | Description                                                            |
|----------------------|------------------------------------------------------------------------|
| `id`                 | Unique participant identifier                                          |
| `age`                | Participant age (years)                                                |
| `baseline_wheat_ige` | Baseline wheat IgE level (kU/L) --- a measure of allergy severity      |
| `vwg_oit`            | Treatment indicator: 1 = VWG OIT, 0 = Placebo                          |
| `max_dose_mg`        | Maximum tolerated wheat protein dose (mg) at the year-1 food challenge |

# Exercise 1: Study design

With your teammate, fill in the following based on the paper and today's lecture:

-   **Treatment indicator** ($Z$):

-   **Treatment group** ($Z = 1$):

-   **Control group** ($Z = 0$):

-   **Outcome** ($Y$):

-   **Covariates** ($X$):

# Exercise 2: Exploratory data analysis

**a.** Create a visualization comparing the distribution of `max_dose_mg` across the two treatment groups.
What do you notice?

```{r}
# add code here
```

*Describe what you see:*

**b.** Compute the mean and standard deviation of `max_dose_mg` for each treatment group.

```{r}
# add code here
```

# Exercise 3: Modeling

**a.** Fit a linear regression model predicting `max_dose_mg` from `vwg_oit`.

```{r}
# add code here
```

**b.** Interpret the slope coefficient for `vwg_oit` in context.
Since this is a randomized experiment, use **causal** language in your interpretation (e.g., "causes," "leads to").

*Write your answer here.*

# Exercise 4: Inference

Choose **one** of the following approaches to assess the evidence for a treatment effect on `max_dose_mg`.
Complete all parts of your chosen option.

------------------------------------------------------------------------

## Option A: Bootstrap confidence interval

**a.** Calculate the observed fit using `specify() |> fit()`.

```{r}
# add code here
```

**b.** Generate 1000 bootstrap samples and fit a model to each one.

```{r}
set.seed(1120)
# add code here
```

**c.** Compute the 95% confidence interval using the percentile method.

```{r}
# add code here
```

**d.** Interpret the confidence interval for `vwg_oit` in context.
Use causal language.

*Write your answer here.*

------------------------------------------------------------------------

## Option B: Hypothesis test

**a.** State the null and alternative hypotheses.

$H_0$:

$H_A$:

**b.** Calculate the observed fit and simulate the null distribution using 1000 permutations.

```{r}
set.seed(24601)
# add code here
```

**c.** Visualize the null distribution and shade the p-value region.

```{r}
# add code here
```

**d.** Calculate the p-value.

```{r}
# add code here
```

**e.** State your conclusion in context.

*Write your answer here.*

------------------------------------------------------------------------

# Exercise 5: Further covariate adjustment

**a.** Fit a model that adds `baseline_wheat_ige` as an additional predictor.
How does the `vwg_oit` estimate compare to the model in Exercise 3?

```{r}
# add code here
```

*Write your answer here.*

**b.** Why might we include `baseline_wheat_ige` in our model even if the two groups are already balanced on this variable?

*Write your answer here.*
