---
title: "AE 20: Wheat Oral Immunotherapy"
subtitle: "Suggested answers"
format:
  html:
    self-contained: true
---

Nowak-Węgrzyn et al. (2019) conducted a multicenter, randomized, double-blind, placebo-controlled clinical trial to evaluate whether vital wheat gluten (VWG) oral immunotherapy (OIT) is effective at increasing wheat tolerance in children and young adults with wheat allergy.
In today's AE, you will analyze a dataset simulated to represent this study using the tools from lecture.

(<https://pubmed.ncbi.nlm.nih.gov/30389226/>)

# Packages

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
library(tidymodels)
```

# Data

```{r}
#| message: false
set.seed(42)
n <- 46
wheat_trial <- tibble(
  id                 = 1:n,
  age                = round(runif(n, 4, 22), 1),
  baseline_wheat_ige = round(rexp(n, rate = 1/30), 1),
  vwg_oit            = c(rep(1, 23), rep(0, 23))
) |>
  mutate(
    max_dose_mg = pmax(0, round(
      150 + 950 * vwg_oit - 3 * baseline_wheat_ige + 12 * age + rnorm(n, sd = 90)
    ))
  )

glimpse(wheat_trial)
```

Variables:

| Variable             | Description                                                            |
|----------------------|------------------------------------------------------------------------|
| `id`                 | Unique participant identifier                                          |
| `age`                | Participant age (years)                                                |
| `baseline_wheat_ige` | Baseline wheat IgE level (kU/L) --- a measure of allergy severity      |
| `vwg_oit`            | Treatment indicator: 1 = VWG OIT, 0 = Placebo                          |
| `max_dose_mg`        | Maximum tolerated wheat protein dose (mg) at the year-1 food challenge |

# Exercise 1: Study design

With your teammate, fill in the following based on the paper and today's lecture:

-   **Treatment indicator** ($Z$): `vwg_oit` --- 1 if the participant received VWG OIT, 0 if placebo

-   **Treatment group** ($Z = 1$): Participants receiving vital wheat gluten (VWG) oral immunotherapy

-   **Control group** ($Z = 0$): Participants receiving a matched placebo

-   **Outcome** ($Y$): `max_dose_mg` --- maximum tolerated wheat protein dose (mg) at the year-1 food challenge

-   **Covariates** ($X$): `age`, `baseline_wheat_ige` (and in the real study: sex, threshold dose at baseline)

# Exercise 2: Exploratory data analysis

**a.** Create a visualization comparing the distribution of `max_dose_mg` across the two treatment groups.
What do you notice?

```{r}
ggplot(
  wheat_trial,
  aes(
    x = factor(vwg_oit, labels = c("Placebo", "VWG OIT")),
    y = max_dose_mg
  )
) +
  geom_boxplot() +
  labs(
    x     = "Treatment group",
    y     = "Max tolerated dose (mg)",
    title = "Max wheat dose by treatment group"
  )
```

*The VWG OIT group has a substantially higher median max tolerated dose than the placebo group, and the distributions show little overlap. This suggests a strong treatment effect.*

**b.** Compute the mean and standard deviation of `max_dose_mg` for each treatment group.

```{r}
wheat_trial |>
  group_by(vwg_oit) |>
  summarise(
    n         = n(),
    mean_dose = mean(max_dose_mg),
    sd_dose   = sd(max_dose_mg)
  )
```

# Exercise 3: Modeling

**a.** Fit a linear regression model predicting `max_dose_mg` from `vwg_oit`.

```{r}
wheat_fit <- linear_reg() |>
  fit(max_dose_mg ~ vwg_oit, data = wheat_trial)

tidy(wheat_fit)
```

**b.** Interpret the slope coefficient for `vwg_oit` in context.
Since this is a randomized experiment, use **causal** language in your interpretation (e.g., "causes," "leads to").

*Receiving VWG OIT **causes** an estimated increase of approximately 950 mg in max tolerated wheat protein dose compared to placebo, on average. Because treatment was randomly assigned, we can attribute this difference to the treatment itself rather than to any confounding variable.*

# Exercise 4: Inference

------------------------------------------------------------------------

## Option A: Bootstrap confidence interval

**a.** Calculate the observed fit using `specify() |> fit()`.

```{r}
observed_fit <- wheat_trial |>
  specify(max_dose_mg ~ vwg_oit) |>
  fit()

observed_fit
```

**b.** Generate 1000 bootstrap samples and fit a model to each one.

```{r}
set.seed(1120)

boot_fits <- wheat_trial |>
  specify(max_dose_mg ~ vwg_oit) |>
  generate(reps = 1000, type = "bootstrap") |>
  fit()

boot_fits
```

**c.** Compute the 95% confidence interval using the percentile method.

```{r}
get_confidence_interval(
  boot_fits,
  point_estimate = observed_fit,
  level          = 0.95,
  type           = "percentile"
)
```

**d.** Interpret the confidence interval for `vwg_oit` in context.
Use causal language.

*We are 95% confident that receiving VWG OIT causes an increase of between approximately \[lower bound\] mg and \[upper bound\] mg in max tolerated wheat protein dose compared to placebo, on average. Because the interval does not include 0, this is consistent with a statistically significant treatment effect.*

------------------------------------------------------------------------

## Option B: Hypothesis test

**a.** State the null and alternative hypotheses.

$H_0$: The slope predicting max tolerated dose from VWG OIT treatment is 0 --- VWG OIT has no effect on max tolerated dose, $\beta_1 = 0$.

$H_A$: The slope is not 0 --- VWG OIT has a significant effect on max tolerated dose, $\beta_1 \neq 0$.

**b.** Calculate the observed fit and simulate the null distribution using 1000 permutations.

```{r}
#| cache: true
#| warning: false
#| message: false
observed_fit <- wheat_trial |>
  specify(max_dose_mg ~ vwg_oit) |>
  fit()

set.seed(24601)
null_dist <- wheat_trial |>
  specify(max_dose_mg ~ vwg_oit) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  fit()
```

**c.** Visualize the null distribution and shade the p-value region.

```{r}
#| warning: false
#| message: false
visualize(null_dist) +
  shade_p_value(obs_stat = observed_fit, direction = "two-sided")
```

**d.** Calculate the p-value.

```{r}
#| warning: false
#| message: false
null_dist |>
  get_p_value(obs_stat = observed_fit, direction = "two-sided")
```

**e.** State your conclusion in context.

*Since* $p < 0.05$, we reject $H_0$.
We have sufficient evidence of a statistically significant treatment effect: VWG OIT **causes** a significant increase in maximum tolerated wheat protein dose compared to placebo.

------------------------------------------------------------------------

# Exercise 5: Further covariate adjustment

**a.** Fit a model that adds `baseline_wheat_ige` as an additional predictor.
How does the `vwg_oit` estimate compare to the model in Exercise 3?

```{r}
wheat_fit_adj <- linear_reg() |>
  fit(max_dose_mg ~ vwg_oit + baseline_wheat_ige, data = wheat_trial)

tidy(wheat_fit_adj)
```

*The `vwg_oit` estimate is very similar to the unadjusted model. This is expected: because treatment was randomly assigned, the two groups should already be balanced on `baseline_wheat_ige`, so including it should not substantially change the treatment effect estimate.*

**b.** Why might we include `baseline_wheat_ige` in our model even if the two groups are already balanced on this variable?

*Even when groups are balanced on a covariate, including it in the model can reduce residual variance and produce more precise (narrower confidence interval) estimates of the treatment effect. In a small trial like this (n = 46), such precision gains can be meaningful.*
